On the effects of partying on next-day sports performance

### 2017-06-12 writing/draft.tex
Polished up the first couple pages a bit more. Added a plot explaining why drinking establishments and population doesn't work.

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
We are interested to see if 
teams exhibit a decline in performance the day following a game in a city 
with active nightlife; part of the question is determining which cities are notorious
for nightlife.

\paragraph{Why do we care} 
Inquiring about this causal relationship is worthwhile for several reasons. The
NBA Commissioner Adam Silver has reported that there is \$400 billion per year spent 
on sports betting in the U.S. alone.\citep{asilver,cnnsilver} Apart from arbitrage opportunities,
knowing how player performance is affected by external factors could inform coaches of a more
optimal play strategy. E.g. if visitation to a city with active nightlife is correlated with
next-day injury, perhaps coaches might bench their star players if the game isn't crucial.

@@

\paragraph{What do we find}
We replicate this analysis for both the National Basketball Association
(NBA) and Major League Baseball (MLB). In particular, we find that visiting
a city with active nightlife the day prior to a game does have a statistically
significant negative effect on a team's likelihood of meeting the point-spread set
by a betting house, for both NBA and MLB. Within the NBA, we analyze specific performance
metrics such as number of points allowed and number of rebounds obtained, and find that
both are negatively effected by previous day visitation to a city with active nightlife.
Lastly, having established a causal model, we realize a profitable betting scheme for MLB.

@@

\subsection{Anecdotal evidence of the effects on sporting performance}
Players openly admit they are interested in nightlife: Matt Barnes recently complained that 
``there's no nightlife in Utah'' when asked about playing Jazz in the second round of NBA playoffs 
in 2017.\citep{cestone}
There are also numerous reports of teams partying to the detriment of the game.
Within the NBA, J.R. Smith's performance metrics substantially improved 
 after being traded from New York to Cleveland; when an interviewer
asked what contributed to this bump in statistics, he infamously replied
that in New York he would have been out partying, but because there's no nightlife in Cleveland, he ends up hitting the gym.\citep{price,ley} Before being traded,
he would score 42 percent of shots taken over the course of his career, but only 34 percent
on Sunday afternoon games in the 2012-13 season, presumably because such games followed a long
weekend of partying.\citep{princeofthecity}

@@

\subsection{Teams}
There has been an attempt to answer the question at the team-level before: i.e. do teams perform
worse after a night on the town, and if so does it depend on which city they 
visited.\citep{ezekowitz} However, the study only
looks at performance in party cities, as opposed to subsequent day performance (i.e. the study does not consider a delayed hangover effect). Moreover, their sample size was much smaller, because the author only considered Sunday afternoon games
from 2006-14. Finally, there was no methodology used behind the selection of party cities;
the authors first tried to define ad-hoc five cities as ``party cities'', and when that did not yield
a significant result they tried estimating fixed-effects for each city. The author does conclude
with weak statistical evidence that teams systematically play worse on Sunday 
afternoons when on the road, regardless of the city.

@@

\section{Defining a Measure for Partying}
A key assumption in our analysis is that next-day opponent is reasonably uncorrelated with
last game location. We plot a heat-map displaying the relationship between last game location
and next game location, and further break this down according to whether the game(s)
were played at home or away.

\begin{figure}
  \hspace{-44pt}
  \includegraphics[scale=0.45]{next_day_opponent}
  \caption{We examine the relationship between last game location and next day opponent in the NBA. Starting from the top left, notice that the most likely scenario across all seasons is that a game will be played at home, and the last game was also at home. This makes sense; 1/2 of all games are played at home, and 1/2 are played away, so clearly each team is most likely to be found playing a home game for any given game. When exactly one of the last game or current game was played away,
we observe that some combinations cease to exist in our data; e.g. we never observe Charlotte Bobcats play Utah Jazz on the road and then fly directly back home for a game. Lastly, notice the patterns exhibited when last game is played on the road, and current game also played on the road. We see that when last game location was in Los Angeles, next day opponent likely to be a west coast team. There appears to be less of a correlation between last game location in New York or Brooklyn and next day opponent, perhaps because players ``gain'' time when flying east-to-west.} \end{figure}

Our takeaway from the figure above is that indeed there is a correlation between last game location
and next day opponent, in particular this is true for west coast teams. We hypothesize that
this is due to practical limitations on flight schedules. The fact that there is a correlation between last game location and next day opponent increases the motivation for conditioning on spread, which accounts may help account for systematic biases to the extent that some of these next-day opponents are particularly (not) skilled.

\paragraph{Partying as a latent variable}
One of the hardest parts of this analysis is that partying is a latent variable.
In order to determine when players partied, one may consider looking at social media data for direct mentions of partying, either by athletes or witnesses. However, due to the intense public scrutiny that athletes face, there is a strong reluctance to disclose their personal lives on social media. There are notable exceptions, such
as Oddel Beckham Jr.'s recent trip to Miami.\citep{bleler}
However, these reports are few and far enough between that picking up statistical evidence would be fruitless.
We develop two approaches: a discrete indicator and a continuous measure of nightlife.

@@

% A *density* plot with overlay instead of two histograms

\begin{sidewaysfigure}
  \hspace{-55pt}
  \includegraphics[scale=0.7,angle=180]{Party_by_Team_Location}
\end{sidewaysfigure}

@@

\subsection{Results}
\subsubsection{NBA: using points-spread data}
We find that after accounting for player fatigue incurred during the last game, jetlag and travel fatigue, recovery time, and home-team effect that both our discrete indicator and continuous measure of party
have a statistically significant effect on likelihood of meeting the spread.
In laymens terms, teams under perform book-maker's expectations after playing in cities 
with active nightlife.
This supports the hypothesis that players enjoy nightlife systematically in certain
cities and that a ``hangover'' effect is felt on the court the next day.

\begin{center}
\begin{tabular}{@{\extracolsep{5pt}}lcc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{2}{c}{\textit{Dependent variable:}} \\  \cline{2-3}  \\[-1.8ex] & \multicolumn{2}{c}{Meet the Spread} \\  \\[-1.8ex] & (1) & (2)\\  \hline \\[-1.8ex]   party discrete & $-$0.513$^{***}$ &  \\    & (0.145) &  \\    & & \\   party continuous &  & $-$0.050$^{**}$ \\    &  & (0.022) \\    & & \\   lag changes in posession & 0.0001 & 0.00004 \\    & (0.002) & (0.002) \\    & & \\   logged travel distance & $-$0.017 & $-$0.015 \\    & (0.026) & (0.026) \\    & & \\   east-west travel direction & $-$0.146 & $-$0.115 \\    & (0.231) & (0.231) \\    & & \\   number hours rest time & 0.001 & 0.001 \\    & (0.001) & (0.001) \\    & & \\   time of game during day & 0.005 & 0.005 \\    & (0.013) & (0.013) \\    & & \\   home team effect & 0.001 & 0.004 \\    & (0.043) & (0.043) \\    & & \\   logged travel distance * east-west & 0.022 & 0.018 \\    & (0.036) & (0.036) \\    & & \\   Constant & $-$0.048 & $-$0.036 \\    & (0.378) & (0.378) \\    & & \\  \hline \\[-1.8ex]  Observations & 9,823 & 9,823 \\  Log Likelihood & $-$6,800.675 & $-$6,804.511 \\  Akaike Inf. Crit. & 13,619.350 & 13,627.020 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular} 
\end{center} 
The features describing jet-lag and recovery time are not significant, but this is to be
expected: such salient information has already been taken into account in setting the spread.
In the next section, we explore which specific areas of gameplay are affected.

@@

\subsubsection{Betting performance}
We also validate our MLB model by seeing how well we can perform against the
market.
\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.82]{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, the 
    circled point represents the worst out of pocket expense incurred.
    We first train a model on historical data then use said model to predict on ``current year''
    data, plotting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet of \$100 if the expected value of the bet is positive after accounting for the bookmaker's cut. E.g. in the     early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have     recouped all of
    our initial investment plus an additional \$11.5k. We are profitable in each season, and mention that 2015 is the ``worst'', but we still come out ahead.}
  \label{bettingperf}
\end{figure}

@@

\begin{figure}   
  \centering
  \hspace{-55pt}
  \includegraphics[scale=0.5]{Party_by_Team_Location_mlb}
  \caption{Continuous measure of nightlife by team location for MLB. What's interesting here is that because New York Yankees are in the Bronx, they show up as having few musicians and sound recording studios. We might wonder if this is a bad thing, since with NBA New York city was deemed a major party city. It turns out that when we define a discrete indicator of party based on New York or Los Angeles, it is not at all significant, and in fact the sign is incorrect. When we use our continuous
 measure of nightlife, its significant, and it helps inform our profitable betting model.
This is good news for our proxy, although begs the question why New York should be treated so differently by MLB players when compared with NBA players. One point worth noting is that we haven't yet made the important distinction that in the MLB, the Los Angeles Angels are in Orange County, for example, which is much less bustling than the NBA's  Los Angeles Lakers who are located in downtown Los Angeles.}
\end{figure}

\begin{figure}   
  \centering
  \hspace{-55pt}
  \includegraphics[scale=0.5]{next_day_opponent_mlb} 
\end{figure}

>>>>

\section{Introduction}
It's not unreasonable to think that in-game sporting performance can be affected 
partly by what takes place off the court. 
We are interested to see if 
teams exhibit a decline in performance the day following a game in a city 
with active nightlife; we call this a ``hangover effect''.
Part of the question is determining a reasonable way to measure
levels of nightlife, and correspondingly which cities are notorious for it; we colloquially
refer to such cities as ``party cities''.

\paragraph{Why do we care} 
Inquiring about this causal relationship is worthwhile for several reasons. The
NBA Commissioner Adam Silver has reported that there is \$400 billion per year spent 
on sports betting in the U.S. alone.\citep{asilver,cnnsilver} Apart from arbitrage opportunities,
knowing how player performance is affected by external factors could inform coaches of a more
optimal play strategy. E.g. if visitation to a city with active nightlife is correlated with
next-day injury, perhaps coaches might bench their star players if the latter game isn't crucial.

@@

\paragraph{What do we find}
We replicate this analysis for both the National Basketball Association
(NBA) and Major League Baseball (MLB). In particular, we find that visiting
a city with active nightlife the day prior to a game does have a statistically
significant negative effect on a team's likelihood of meeting the point-spread set
by a betting house, for both NBA and MLB. Within the NBA, we analyze specific performance
metrics such as number of points allowed and number of rebounds obtained, and find that both are negatively effected by previous day visitation to a city with active nightlife. Having established a causal model, we realize a profitable betting scheme for MLB. We also perform a robustness check and see that if players rest more than
 24 hours after visiting a ``party-city'', the hangover effect dissipates.

@@

\subsection{Anecdotal evidence of the effects on sporting performance}
Professional players openly admit they are interested in nightlife: Matt Barnes recently complained that 
``there's no nightlife in Utah'' when asked about playing Jazz in the second round of NBA playoffs 
in 2017.\citep{cestone}
There are also numerous reports of teams partying to the detriment of the game.
Within the NBA, J.R. Smith's performance metrics substantially improved 
 after being traded from New York to Cleveland; when an interviewer
asked what contributed to this bump in performance, he infamously replied
that in New York he would have been out partying, but because there's no nightlife in Cleveland he ends up hitting the gym.\citep{price,ley} Before being traded, J.R. Smith's career-to-date success rate of shots taken was 42 percent, but somehow he only managed 
34 percent on Sunday afternoon games in the 2012-13 season.
Presumably, this was because such games followed a long weekend of partying.\citep{princeofthecity}

@@

\subsection{Teams}
There has been an attempt to answer the question at the team-level before: i.e. do teams perform
worse after a night on the town, and if so does it depend on which city they 
visited.\citep{ezekowitz} However, the study only
looks at performance in party cities, as opposed to subsequent day performance (i.e. the study does not consider a delayed hangover effect). Moreover, their sample size was much smaller, because the author only considered Sunday afternoon games
from 2006-14. Finally, there was no methodology used behind the selection of party cities:
the authors first tried to define ad-hoc five ``party cities'', and when that did not yield
a significant result they tried estimating fixed-effects for each city. The author does conclude
with weak statistical evidence that teams systematically play worse on Sunday 
afternoons when on the road, regardless of the city.

@@

\section{Defining a Measure for Partying}

\paragraph{On the correlation between next day opponent and last game location}
A key assumption in our analysis is that next-day opponent is reasonably uncorrelated with
last game location. We plot a heat-map displaying the relationship between last game location
and next game location, and further break this down according to whether the game(s)
were played at home or away.

\begin{figure}
  \centering
  \label{fig: next day opponent}
  \includegraphics[scale=0.45]{next_day_opponent}
  \caption{We examine the relationship between last game location and next day opponent in the NBA. Starting from the top left, notice that the most likely scenario across all seasons is that a game will be played at home, and the last game was also at home. This makes sense; 1/2 of all games are played at home, and 1/2 are played away, so clearly each team is most likely to be found playing a home game for any given game. When exactly one of the last game or current game was played away,
we observe that some combinations cease to exist in our data; e.g. we never observe Charlotte Bobcats play Utah Jazz on the road and then fly directly back home for a game. Lastly, notice the patterns that emerge when both the last game and current game are played on the road. E.g. we see that when last game location was in Los Angeles, next day opponent likely to be a west coast team. There appears to be less of a correlation between last game location in New York or Brooklyn and next day opponent, perhaps because flying from east-to-west
is less burdensome as it relates to timezone changes.} 
\end{figure}

Our takeaway from the figure on page \ref{fig: next day opponent} is that indeed there is some correlation between last game location
and next day opponent, and in particular this is true for west coast teams. We hypothesize that
this is due to practical limitations on flight schedules. E.g. if a team is unwilling to take a red-eye
flight, then there is little hope to play an evening west coast game and make it to the east coast for a game the very next day;
the same is \emph{not true} when flying from east to west, in which case a morning flight is reasonable.
The fact that there is some correlation between last game location and next day opponent, even if only in west coast cities, increases the motivation for conditioning on spread, which accounts may help account for systematic biases to the extent that some of these next-day opponents are particularly (not) skilled.

\paragraph{Partying as a latent variable}
One of the hardest parts of this analysis is that partying is a latent variable.
In order to determine when players partied, one may consider looking at social media data for direct mentions of partying, either by athletes or witnesses. However, due to the intense public scrutiny that athletes face, there is a strong reluctance to disclose their personal lives on social media. There are notable exceptions, such
as Oddel Beckham Jr.'s recent trip to Miami.\citep{bleler}
However, these reports are few and far enough between that picking up statistical evidence would be fruitless.
Instead, we develop two approaches which do not rely on this data: a discrete indicator 
based on player interviews and also a continuous measure of nightlife based on local community patterns.

@@

\begin{figure}
  \centering
  \label{fig: party by team loc}
  \includegraphics[scale=0.6,angle=270]{Party_by_Team_Location}
  \caption{We plot our continuous measure of nightlife for each city within the NBA. We present the analagous plot for MLB in the appendix.\ref{fig: party by team location mlb}}
\end{figure}

@@

\subsection{Results}
\subsubsection{NBA: using points-spread data}
We find that after accounting for player fatigue incurred during the last game, jetlag and travel fatigue, recovery time, and home-team effect that both our discrete indicator and continuous measure of party
have a statistically significant effect on likelihood of meeting the spread.
In laymens terms, teams under perform book-maker's expectations after playing in cities 
with active nightlife.

\begin{center}
\begin{tabular}{@{\extracolsep{5pt}}lcc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{2}{c}{\textit{Dependent variable:}} \\  \cline{2-3}  \\[-1.8ex] & \multicolumn{2}{c}{Meet the Spread} \\ \hline \\[-1.8ex]   party discrete & $-$0.513$^{***}$ &  \\    & (0.145) &  \\    & & \\   party continuous &  & $-$0.050$^{**}$ \\    &  & (0.022) \\    & & \\   lag changes in posession & 0.0001 & 0.00004 \\    & (0.002) & (0.002) \\    & & \\   logged travel distance & $-$0.017 & $-$0.015 \\    & (0.026) & (0.026) \\    & & \\   east-west travel direction & $-$0.146 & $-$0.115 \\    & (0.231) & (0.231) \\    & & \\   number hours rest time & 0.001 & 0.001 \\    & (0.001) & (0.001) \\    & & \\   time of game during day & 0.005 & 0.005 \\    & (0.013) & (0.013) \\    & & \\   home team effect & 0.001 & 0.004 \\    & (0.043) & (0.043) \\    & & \\   logged travel distance * east-west & 0.022 & 0.018 \\    & (0.036) & (0.036) \\    & & \\   Constant & $-$0.048 & $-$0.036 \\    & (0.378) & (0.378) \\ \hline \\[-1.8ex]  Observations & 9,823 & 9,823 \\  Log Likelihood & $-$6,800.675 & $-$6,804.511 \\  Akaike Inf. Crit. & 13,619.350 & 13,627.020 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01}
\end{tabular} 
\end{center}

This model supports the hypothesis that players enjoy nightlife systematically in certain
cities and that a ``hangover'' effect is felt on the court the next day.
The features describing jet-lag and recovery time are not significant, but this is to be
expected: such salient information has already been taken into account in setting the spread.
In the next section, we explore which specific areas of gameplay are affected.

@@

\subsubsection{Betting performance}
We also validate our MLB model by seeing how well we can perform against the
market.
\begin{figure}[!h]
  \centering
  \label{mlb bets by season}
  \includegraphics[scale=0.82]{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, the 
    circled point represents the worst out of pocket expense incurred.
    We first train a model on historical data then use said model to predict on ``current year''
    data, plotting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet of \$100 if the expected value of the bet is positive after accounting for the bookmaker's cut. E.g. in the     early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have     recouped all of
    our initial investment plus an additional \$11.5k. We are profitable in each season, and mention that 2015 is the ``worst'', but we still come out ahead.}
  \label{bettingperf}
\end{figure}

@@

\begin{figure}   
  \centering
  \hspace{-55pt}
  \label{fig: party by team location mlb}
  \includegraphics[scale=0.5]{Party_by_Team_Location_mlb}
  \caption{Continuous measure of nightlife by team location for MLB. What's interesting here is that because New York Yankees are in the Bronx, they show up as having few musicians and sound recording studios. We might wonder if this is a bad thing, since with NBA New York city was deemed a major party city. It turns out that when we define a discrete indicator of party based on New York or Los Angeles, it is not at all significant, and in fact the sign is incorrect. When we use our continuous
 measure of nightlife, its significant, and it helps inform our profitable betting model.
This is good news for our proxy, although begs the question why New York should be treated so differently by MLB players when compared with NBA players. One point worth noting is that we haven't yet made the important distinction that in the MLB, the Los Angeles Angels are in Orange County, for example, which is much less bustling than the NBA's  Los Angeles Lakers who are located in downtown Los Angeles.}
\end{figure}

\begin{figure}   
  \centering
  \hspace{-55pt}
  \label{fig: next day opponent mlb}
  \includegraphics[scale=0.5]{next_day_opponent_mlb} 
\end{figure}

\begin{sidewaysfigure}   
  \centering
  \label{fig: why drinks or pop dont work}
  \includegraphics[scale=0.5]{why_drinks_or_population_dont_work} 
  \caption{We remark why using a statistic such as number of drinking establishments or 
    population size doesn't work. Although both of these are linearly correlated with musicians, 
    and even in spite of the extremal observations being ranked consistently across each statistic, 
    i.e. Charlotte Bobcats at the bottom and Los Angeles at the top,  
    the problem is that population and number of drinking establishments have sinusoidal 
    relationships with respect to the rank ordering imposed by the musicians index 
    that do not properly reflect variation in nightlife between cities. For example, both 
    incorrectly rank Detroit Michigan as having a high nightlife index. Another example, population incorrectly places New Orleans, Louisiana as having a low nightlife index, when in reality it is famous for its vibrant city life.}
\end{sidewaysfigure}
